Generalization Errors of Hard Parameter Sharing: High-dimensional Asymptotics and Implications

### 2020-09-30 arxiv/proof_examples.tex
revise

\paragraph{Proofs for two tasks.}  We now give the proof of Proposition \ref{cor_MTL_loss}. As in the proof of Theorem \ref{thm_many_tasks}, we first derive a bias-variance decomposition of the prediction loss $L(\hat \eta_2^{\MTL})$. Then we can calculate the bias term and the variance term using Theorem \ref{thm_main_RMT}.

@@

		f(A, B) =   \bignorm{X^{(1)} B A_1 - Y^{(1)}}^2+ \bignorm{X^{(2)} B A_2 - Y^{(2)}}^2, \label{eq_mtl_2task_cov}

@@

&+ A^2[(A-1)^2\kappa^2 + (A^2+1)d^2/2] \cdot p^{-1}\tr\left[  (X^{(1)})^\top X^{(1)} \hat \Sigma^{-1}(X^{(2)})^\top X^{(2)} \hat \Sigma^{-1} (X^{(1)})^\top X^{(1)}\right] +\sigma^2(n_1+n_2-p),

@@

where we used that the entries of $\beta_1-A\beta_2\in \R^p$ are i.i.d. random variables with mean zero and variance $p^{-1}[(A-1)^2\kappa^2 + (A^2+1)d^2/2]$.

@@

Then we calculate the predication loss ...

% \todo{switch the $a_3$ and $a_4$}

For the isotropic model in Section \ref{sec_similarity}, we actually have an easier and sharper bound than Lemma \ref{thm_model_shift}. %as follows.
\begin{lemma}\label{prop_model_shift_tight}
		In the setting of Lemma \ref{thm_model_shift}, assume that $\Sigma_1 =\id$,
		$\beta_2$ is a random vector that has i.i.d. entries with mean $0$, variance $\kappa^2$ and finite moments up to any order, and $(\beta_1-\beta_2)$ is a random vector that is independent of $\beta_2$ and has i.i.d. entries with mean $0$, variance $d^2$ and finite moments up to any order. Denote
		$\Delta^\star_{\bias} := \bigbrace{(1 - \hat{v})^2 \kappa^2 + d^2} \bigtr{\Pi(\hat v)}$, where $\Pi$ was defined in \eqref{defnpihat}.	Then we have with high probability,
		\begin{align*}
			\te(\hat{\beta}_t^{\MTL}) \le \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \ge  (\al_+^2(\rho_1)+\oo(1))\cdot  \Delta^\star_{\bias} \ , \\
			\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \le (\al_-^2(\rho_1)-\oo(1))\cdot  \Delta^\star_{\bias}\ ,
		\end{align*}
		where $\Delta_{\vari}$ was defined in \eqref{Deltavarv}.
\end{lemma}

We show that for the isotropic model, the optimum of $\hat v$ is very close to 1.
This allows us to simplify the expression of $\te(\hat{\beta}_2^{\MTL})$ significantly.
%Our main observation is that the optimum of $\val(v)$ is approximately $1$.This allows us to simplify the bias and variance bounds significantly.
%Furthermore, under certain choice of parameters, we can simplify  
More precisely, we have the following lemma.
Later on, we will combine it with Lemma \ref{prop_model_shift_tight} together to analyze the bias-variance tradeoff.

@@

%For the rest of this section, we choose parameters that satisfy the following relations: \be\label{choiceofpara}
%pd^2 \sim \sigma^2 \sim 1,\quad p^{-1+c_0} \sigma^2 \le \kappa^2  \le p^{-\e_0-c_0}\sigma^2 ,
%\ee
%for some small constant $c_0>0$. 
Next we prove the following estimate on the optimizer $\hat v$: with high probability,
%\begin{lemma}
%For the isotropic model, we have
\be\label{hatv_add1}
|\hat v - 1|= \OO\left(\cal E\right), \quad \cal E:=\frac{d^2}{\kappa^2} + \frac{\sigma^2}{p\kappa^2} + p^{-1/2 + \e_0 /2+ 2\e}.
\ee
%\end{lemma}
%\begin{proof}
 In fact, from the proof of Claim \ref{lem_hat_v} above, one can check that if $C\cal E \le |w-1| \le 2C\cal E$ for a large enough constant $C>1$, then $|h'(w)|\gtrsim \sigma^2$. Moreover, under \eqref{choiceofpara} we have
$$h(w) =\OO(\sigma^2),\quad \text{for}\quad   |w-1|\le 2C\cal E.$$
Thus we obtain that for $|w-1|\ge 2C\cal E$,
$$\left|h(w) - h(\hat w)\right|\ge |h(w)-\min\{h(w_1),h(w_2)\}|\gtrsim \sigma^2 \cal E \gtrsim \cal E \cdot h(\hat w),$$
which leads to $g(w) > g(\hat w)$ with high probability by \eqref{boundv-w}. Thus $w$ cannot be a minimizer of $g(v)$, and we must have $|\hat v - 1|\le 2C\cal E$. %Together with \eqref{hatw_add1}, we conclude \eqref{hatv_add1}.
%\end{proof}

Inserting \eqref{hatv_add1} into \eqref{eq_te_mtl_2task} and applying Lemma  \ref{largedeviation} to $(\beta_1-\hat v\beta_s)$, we get that w.h.p.,
\begin{align}
\te(\hat{\beta}_2^{\MTL})&=(1+\OO(\cal E))\cdot \left[d^2 + \OO\left(\cal E^2 \kappa^2\right)\right] \tr\left[(X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right]\nonumber\\
&+(1+\OO(\cal E))\cdot \sigma^2  \bigtr{(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} }. \label{1Liso}
\end{align}
In order to study the phenomenon of bias-variance trade-off, we need the bias term with $d^2$ and the variance term with $\sigma^2$ to be of the same order. Using Fact \eqref{lem_minv}, we have that
$$\tr\left[(X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right] \sim p,\quad \bigtr{(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} } \sim 1.$$
Hence we need to choose that $p d^2 \sim \sigma^2$. On the other hand, we want the error term $\cal E^2 \kappa^2$ to be much smaller than $d^2$, which leads to the condition $p^{-1+\e_0+4\e}\kappa^2  \ll d^2 \ll \kappa^2$. The above considerations lead to the choices of parameters in \eqref{choiceofpara}, under which we can simplify \eqref{1Liso} to \eqref{simple1}.
%\begin{align}
%\te(\hat{\beta}_2^{\MTL})&=(1+\OO(n^{-\e}))\cdot d^2 \tr\left[(X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right] \nonumber \\
%&+(1+\OO(n^{-\e}))\cdot \sigma^2  \bigtr{(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} }  \label{simple1}
%\end{align}
%w.h.p. for some constant $\e>0$.
\end{proof}

%\subsection{Covariate Shift}\label{app_proof_33}

>>>>

\subsubsection{Proof for Sample Imbalance}  
We now give the proof of Proposition \ref{cor_MTL_loss}. As in the proof of Theorem \ref{thm_many_tasks}, our proof is divided into the following steps: 
\begin{itemize}
\item simplifying the optimization objective using concentration estimates;

\item finding the minimizer of the optimization objective;

\item deriving a bias-variance decomposition of the prediction loss $L(\hat \beta_2^{\MTL})$;

\item calculating the bias term and the variance term using Theorem \ref{thm_main_RMT}. 
\end{itemize}

\paragraph{Simplifying the optimization objective.} 

@@

		f(A_1,A_2, B) =   \bignorm{X^{(1)} B A_1 - Y^{(1)}}^2+ \bignorm{X^{(2)} B A_2 - Y^{(2)}}^2, \label{eq_mtl_2task_cov}

@@

&+ A^2[(A-1)^2\kappa^2 + (A^2+1)d^2/2] \cdot p^{-1}\tr\left[  (X^{(1)})^\top X^{(1)} \hat \Sigma^{-1}(X^{(2)})^\top X^{(2)} \hat \Sigma^{-1} (X^{(1)})^\top X^{(1)}\right] +\sigma^2(n_1+n_2-p)\\
&=[(A-1)^2\kappa^2 + (A^2+1)d^2/2] \cdot p^{-1} \tr\left[ (X^{(1)})^\top X^{(1)} \hat \Sigma^{-1}(X^{(2)})^\top X^{(2)}\right]+\sigma^2(n_1+n_2-p),

@@

where in the first step we used that the entries of $\beta_1-A\beta_2\in \R^p$ are i.i.d. random variables with mean zero and variance $p^{-1}[(A-1)^2\kappa^2 + (A^2+1)d^2/2]$, and in the second step we used  %Using $(X^{(2)})^\top X^{(2)}=\hat\Sigma- A^2 (X^{(1)})^\top X^{(1)}$, we can further simplify that
 &\tr\left[  (X^{(2)})^\top X^{(2)} \hat \Sigma^{-1}(X^{(1)})^\top X^{(1)} \hat \Sigma^{-1} (X^{(2)})^\top X^{(2)}\right]+ A^2 \tr\left[  (X^{(1)})^\top X^{(1)} \hat \Sigma^{-1}(X^{(2)})^\top X^{(2)} \hat \Sigma^{-1} (X^{(1)})^\top X^{(1)}\right] \\
 &=\tr\left[  \hat \Sigma^{-1}(X^{(1)})^\top X^{(1)} \hat \Sigma^{-1} \left( \hat\Sigma- A^2 (X^{(1)})^\top X^{(1)} \right)^2\right]+ A^2 \tr\left[  (X^{(1)})^\top X^{(1)} \hat \Sigma^{-1}\left( \hat\Sigma- A^2 (X^{(1)})^\top X^{(1)} \right) \hat \Sigma^{-1} (X^{(1)})^\top X^{(1)}\right] \\
 &=\tr\left[  (X^{(1)})^\top X^{(1)}\right] - A^2\tr\left[ (X^{(1)})^\top X^{(1)} \hat \Sigma^{-1}(X^{(1)})^\top X^{(1)}\right] = \tr\left[ (X^{(1)})^\top X^{(1)} \hat \Sigma^{-1}(X^{(2)})^\top X^{(2)}\right] .

@@

 \paragraph{Minimizer of the optimization objective.} We show that for the isotropic model, the optimum of $\hat v$ is very close to 1.
This allows us to simplify the expression of $\te(\hat{\beta}_2^{\MTL})$ significantly.
%Our main observation is that the optimum of $\val(v)$ is approximately $1$.This allows us to simplify the bias and variance bounds significantly.
%Furthermore, under certain choice of parameters, we can simplify  
More precisely, we have the following lemma.
Later on, we will combine it with Lemma \ref{prop_model_shift_tight} together to analyze the bias-variance tradeoff.

@@

%For the rest of this section, we choose parameters that satisfy the following relations: \be\label{choiceofpara}
%pd^2 \sim \sigma^2 \sim 1,\quad p^{-1+c_0} \sigma^2 \le \kappa^2  \le p^{-\e_0-c_0}\sigma^2 ,
%\ee
%for some small constant $c_0>0$. 
Next we prove the following estimate on the optimizer $\hat v$: with high probability,
%\begin{lemma}
%For the isotropic model, we have
\be\label{hatv_add1}
|\hat v - 1|= \OO\left(\cal E\right), \quad \cal E:=\frac{d^2}{\kappa^2} + \frac{\sigma^2}{p\kappa^2} + p^{-1/2 + \e_0 /2+ 2\e}.
\ee
%\end{lemma}
%\begin{proof}
 In fact, from the proof of Claim \ref{lem_hat_v} above, one can check that if $C\cal E \le |w-1| \le 2C\cal E$ for a large enough constant $C>1$, then $|h'(w)|\gtrsim \sigma^2$. Moreover, under \eqref{choiceofpara} we have
$$h(w) =\OO(\sigma^2),\quad \text{for}\quad   |w-1|\le 2C\cal E.$$
Thus we obtain that for $|w-1|\ge 2C\cal E$,
$$\left|h(w) - h(\hat w)\right|\ge |h(w)-\min\{h(w_1),h(w_2)\}|\gtrsim \sigma^2 \cal E \gtrsim \cal E \cdot h(\hat w),$$
which leads to $g(w) > g(\hat w)$ with high probability by \eqref{boundv-w}. Thus $w$ cannot be a minimizer of $g(v)$, and we must have $|\hat v - 1|\le 2C\cal E$. %Together with \eqref{hatw_add1}, we conclude \eqref{hatv_add1}.
%\end{proof}
\end{proof}

\paragraph{Calculating the prediction loss.} Then we calculate the predication loss ...

 Inserting \eqref{hatv_add1} into \eqref{eq_te_mtl_2task} and applying Lemma  \ref{largedeviation} to $(\beta_1-\hat v\beta_s)$, we get that w.h.p.,
\begin{align}
\te(\hat{\beta}_2^{\MTL})&=(1+\OO(\cal E))\cdot \left[d^2 + \OO\left(\cal E^2 \kappa^2\right)\right] \tr\left[(X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right]\nonumber\\
&+(1+\OO(\cal E))\cdot \sigma^2  \bigtr{(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} }. \label{1Liso}
\end{align}
In order to study the phenomenon of bias-variance trade-off, we need the bias term with $d^2$ and the variance term with $\sigma^2$ to be of the same order. Using Fact \eqref{lem_minv}, we have that
$$\tr\left[(X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right] \sim p,\quad \bigtr{(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} } \sim 1.$$
Hence we need to choose that $p d^2 \sim \sigma^2$. On the other hand, we want the error term $\cal E^2 \kappa^2$ to be much smaller than $d^2$, which leads to the condition $p^{-1+\e_0+4\e}\kappa^2  \ll d^2 \ll \kappa^2$. The above considerations lead to the choices of parameters in \eqref{choiceofpara}, under which we can simplify \eqref{1Liso} to \eqref{simple1}.
%\begin{align}
%\te(\hat{\beta}_2^{\MTL})&=(1+\OO(n^{-\e}))\cdot d^2 \tr\left[(X_1^{\top}X_1 +X_2^{\top}X_2)^{-2} (X_1^{\top}X_1)^2\right] \nonumber \\
%&+(1+\OO(n^{-\e}))\cdot \sigma^2  \bigtr{(X_1^{\top}X_1  + X_2^{\top}X_2)^{-1} }  \label{simple1}
%\end{align}
%w.h.p. for some constant $\e>0$.

% \todo{switch the $a_3$ and $a_4$}
\paragraph{Calculating the variance and bias terms.} For the isotropic model in Section \ref{sec_similarity}, we actually have an easier and sharper bound than Lemma \ref{thm_model_shift}. %as follows.
\begin{lemma}\label{prop_model_shift_tight}
		In the setting of Lemma \ref{thm_model_shift}, assume that $\Sigma_1 =\id$,
		$\beta_2$ is a random vector that has i.i.d. entries with mean $0$, variance $\kappa^2$ and finite moments up to any order, and $(\beta_1-\beta_2)$ is a random vector that is independent of $\beta_2$ and has i.i.d. entries with mean $0$, variance $d^2$ and finite moments up to any order. Denote
		$\Delta^\star_{\bias} := \bigbrace{(1 - \hat{v})^2 \kappa^2 + d^2} \bigtr{\Pi(\hat v)}$, where $\Pi$ was defined in \eqref{defnpihat}.	Then we have with high probability,
		\begin{align*}
			\te(\hat{\beta}_t^{\MTL}) \le \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \ge  (\al_+^2(\rho_1)+\oo(1))\cdot  \Delta^\star_{\bias} \ , \\
			\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \le (\al_-^2(\rho_1)-\oo(1))\cdot  \Delta^\star_{\bias}\ ,
		\end{align*}
		where $\Delta_{\vari}$ was defined in \eqref{Deltavarv}.
\end{lemma}

\subsection{Covariate Shift}\label{app_proof_33}
